Prospects of detecting gamma-ray emission from galaxy clusters: cosmic rays and dark matter annihilations

### 2011-10-13 paper.tex
small changes

Figure~\ref{fig14} compares \Fermi upper limits on the gamma-ray flux
with predictions of DM annihilation fluxes in the LP and BM
models. \C{We assume a boost factor due to substructures that has a
  constant contribution per decade in substructure mass and has a mass
  spectrum extending down to Earth masses for our LP and BM DM
  annihilation models. In the LP models we additionally assume a
  cluster-mass dependent boost due to SFE. The gamma-ray fluxes are
  calculated for those clusters where 18 months \Fermi-LAT upper
  limits are derived (see \cite{2010ApJ...717L..71A}).} \del{To obtain
  the gamma-ray fluxes, we scale the virial mass of clusters in the
  extended HIFLUGCS catalogue \cite{2007A&A...466..805C} with our
  scaling relation of Eq.~(\ref{eq:DM_scaling}). In those, we assume a
  boost factor due to substructures that has a constant contribution
  per decade in substructure mass and has a mass spectrum extending
  down to Earth masses for our LP and BM DM annihilation models. In
  the LP models we additionally assume a cluster-mass dependent boost
  due to SFE.}  \del{\Fermi} \C{These} limits rule out the LP models
in their present form with the mentioned assumptions in 28 clusters,
and limit the boost from SFE to less than 5 in M49 and
Fornax. Assuming universality, this limits the Sommerfeld boost in the
MW to less than 3 (see Fig.~\ref{fig:boost_const}). The flux level of
the \Fermi limits are an indirect measure of the ambient background
flux in the gamma-ray sky and/or the presence of strong point sources
such as AGN in the immediate vicinity of the cluster position. Hence,
the ratio of the \Fermi upper limits to the predicted annihilation
fluxes, $F_{\mathrm{UL}}/F_{\mathrm{DM}}$, is a good indication of the
best cluster candidates for indirect DM experiments. We identify
Fornax, M49, NGC4636, and Virgo to be prime candidates. Note however,
that Virgo extends of $12^\circ$ over the sky which implies a lower
surface brightness and lower signal-to-noise. \Fermi limits on
individual clusters are expected to improve as $\sqrt{T/1.5
  \mathrm{yr}}$, where $T$ is the total elapsed time of the \Fermi
mission. We emphasize that the very inhomogeneous distribution of
$F_{\mathrm{UL}} / F_{\mathrm{DM}}$ makes it unlikely to dramatically
improve the limits through a stacking analysis as the weak signal will
dilute the average constraints.

In Figure~\ref{fig21}, we show the DM annihilation flux predictions
\del{of the entire sample} \C{of all the clusters in the extended
  HIFLUGCS catalogue \cite{2007A&A...466..805C}. We find M49 and
  Fornax to be that brightest clusters in the entire sample. While the
  LP model predicts that almost all clusters in the sample should be
  observed by \Fermi after two years of data taking, the fluxes from
  the BM models are too low to be detected in the near future without
  improved modelling.} Assuming the projected point source
sensitivity ($5\sigma$, 50h) of the future Cherenkov telescope array
(CTA) of $F_\gamma(>E_\gamma) = (4\times10^{-11}, 2\times10^{-12},
2\times10^{-14})\,\rmn{ph}\,\rmn{cm}^{-2}\,\rmn{s}^{-1}$ at energies
of $E_\gamma=(10~\rmn{GeV}, 100~\rmn{GeV}, 1~\rmn{TeV})$
\cite{Doro:2009qs}, we note that it will be very challenging to detect
the DM annihilation signal from clusters without a Sommerfeld boost by
Cherenkov telescopes. This is because the boost from substructures is
extended while the sensitivity of Cherenkov telescopes scales
approximately linearly with source extension. Hence the fluxes quoted
in Fig.~\ref{fig21} will have to be compared to a sensitivity that is
scaled down by the ratio of cluster radius to angular resolution of
$0.1^\circ$ assuming current background subtraction techniques of
Cherenkov telescopes. This important finding should encourage the
development of new methods to overcome the degradation of sensitivity
for diffuse and very extended sources. Such a break-through would be
needed to probe and potentially detect the presented class of BM
models with a large investment of observational time.

>>>>

Figure~\ref{fig14} compares \Fermi upper limits on the gamma-ray flux
with predictions of DM annihilation fluxes in the LP and BM
models. \C{We assume a boost factor due to substructures that has a
  constant contribution per decade in substructure mass and has a mass
  spectrum extending down to Earth masses for our LP and BM DM
  annihilation models.}  \del{In the LP models we additionally assume
  a cluster-mass dependent boost due to SFE.}\C{The gamma-ray fluxes
  are calculated for those clusters where 18 months \Fermi-LAT upper
  limits are derived (see \cite{2010ApJ...717L..71A}).} \del{To obtain
  the gamma-ray fluxes, we scale the virial mass of clusters in the
  extended HIFLUGCS catalogue \cite{2007A&A...466..805C} with our
  scaling relation of Eq.~(\ref{eq:DM_scaling}). In those, we assume a
  boost factor due to substructures that has a constant contribution
  per decade in substructure mass and has a mass spectrum extending
  down to Earth masses for our LP and BM DM annihilation models. In
  the LP models we additionally assume a cluster-mass dependent boost
  due to SFE.}  \del{\Fermi} \C{These} limits rule out the LP models
in their present form with the mentioned assumptions in 28 clusters,
and limit the boost from SFE to less than 5 in M49 and
Fornax. Assuming universality, this limits the Sommerfeld boost in the
MW to less than 3 (see Fig.~\ref{fig:boost_const}). The flux level of
the \Fermi limits are an indirect measure of the ambient background
flux in the gamma-ray sky and/or the presence of strong point sources
such as AGN in the immediate vicinity of the cluster position. Hence,
the ratio of the \Fermi upper limits to the predicted annihilation
fluxes, $F_{\mathrm{UL}}/F_{\mathrm{DM}}$, is a good indication of the
best cluster candidates for indirect DM experiments. We identify
Fornax, M49, NGC4636, and Virgo to be prime candidates. Note however,
that Virgo extends $12^\circ$ over the sky which implies a lower
surface brightness and lower signal-to-noise. \Fermi limits on
individual clusters are expected to improve as $\sqrt{T/1.5
  \mathrm{yr}}$, where $T$ is the total elapsed time of the \Fermi
mission. \C{We emphasize that the very inhomogeneous distribution of
  $F_{\mathrm{UL}} / F_{\mathrm{DM}}$ makes it unlikely to
  dramatically improve the limits through a likelihood stacking
  analysis as there are only a few clusters with comparably good
  $F_{\mathrm{UL}}/F_{\mathrm{DM}}$ values as well as

In Figure~\ref{fig21}, we show the DM annihilation flux predictions
\del{of the entire sample}\C{of all the clusters in the extended
  HIFLUGCS catalogue \cite{2007A&A...466..805C}. We find M49 and
  Fornax to be that brightest clusters in the entire sample. While the
  LP model predicts that almost all clusters in the sample should be
  observed by \Fermi after two years of data taking, the fluxes from
  the BM models are too low to be detected in the near future without
  improved modelling.} Assuming the projected point source
sensitivity ($5\sigma$, 50h) of the future Cherenkov telescope array
(CTA) of $F_\gamma(>E_\gamma) = (4\times10^{-11}, 2\times10^{-12},
2\times10^{-14})\,\rmn{ph}\,\rmn{cm}^{-2}\,\rmn{s}^{-1}$ at energies
of $E_\gamma=(10~\rmn{GeV}, 100~\rmn{GeV}, 1~\rmn{TeV})$
\cite{Doro:2009qs}, we note that it will be very challenging to detect
the DM annihilation signal from clusters without a Sommerfeld boost by
Cherenkov telescopes. This is because the boost from substructures is
extended while the sensitivity of Cherenkov telescopes scales
approximately linearly with source extension. Hence the fluxes quoted
in Fig.~\ref{fig21} will have to be compared to a sensitivity that is
scaled down by the ratio of cluster radius to angular resolution of
$0.1^\circ$ assuming current background subtraction techniques of
Cherenkov telescopes. This important finding should encourage the
development of new methods to overcome the degradation of sensitivity
for diffuse and very extended sources. Such a break-through would be
needed to probe and potentially detect the presented class of BM
models with a large investment of observational time.
